### 2016-02-13 Paper_v5.tex

\subsection*{Dynamic optimization with particle swarms (DOPS).}
DOPS is a novel meta-heuristic which combines multi-swarm particle swarm methods with dynamically dimensioned search (Fig. \ref{fig-algorithm}).
The goal of DOPS is to estimate optimal or near optimal parameter vectors for high-dimensional biological models within a specified number of function evaluations.
Toward this objective, DOPS begins by using a particle swarm search and then dynamically switches, using an adaptive switching criteria, to the DDS search phase.
We began the particle swarm phase by randomly initializing a swarm of $\mathcal{K}$-dimensional particles (represented as $z_{i}$), wherein each particle corresponded to a $\mathcal{K}$-dimensional parameter vector.
After initialization, particles were randomly partitioned into $k$ equal sized sub-swarms $\mathcal{S}_{1},\hdots,\mathcal{S}_{k}$.
Thereafter within each sub-swarm $\mathcal{S}_{k}$, particles were updated according to the rule:
\begin{equation}\label{eqn:update-rule}
	\mathbf{z}_{i,j} = \theta_{1,j-1}\mathbf{z}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{z}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G}_{k} - \mathbf{z}_{i,j-1}\right)
\end{equation}
where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ were adjustable parameters, $\mathcal{L}_{i}$ denotes the best solution found by particle $i$ within sub-swarm $k$ for function evaluation $1\rightarrow j-1$, and
$\mathcal{G}_{k}$ denotes the best solution found over all particles within sub-swarm $\mathcal{S}_{k}$.
The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model parameters ($\mathcal{K}\times{1}$).
Equation (3) is similar to the general particle swarm update rule, however, it does not contain velocity terms.
In DOPS the parameter $\theta_{1,j-1}$ depends upon the function evaluations and is updated according to:
\begin{eqnarray}
	\mathbf \theta_{1,j}&=&\frac{(\mathcal{N}-{j})*(\mathbf{w}_{max}-\mathbf{w}_{min}))}{(\mathcal{N}-{1})} + \mathbf{w}_{min}
\end{eqnarray}
where $\mathcal{N}$ represents the total number of function evaluations, $\mathbf{w}_{max}$ and $\mathbf{w}_{min}$ are the maximum and minimum inertia weights, respectively.
While updating the particles, we made sure all dimensions of the solution represented by the particle were within bounds using a set of reflection boundary conditions:

@@

After every $\mathcal{M}$ function evaluations, particles were randomly redistributed to a new sub-swarm, and updated according to Eqn. \eqref{eqn:update-rule}.
This process continued till $\mathcal{F}*\mathbf{N}$ number of functions evaluations, where $\mathcal{F}$ is the fraction of evaluations with the multi-swarms.
The fraction of evaluations within the swam phase is based on the drop in error.
If the error does not drop by 1\% for a certain number of evaluations the search within swarm phase is terminated.
At the end of these function evaluations, we froze all the solutions represented by various particles and chose the particle with best solution among $\mathcal{G}_{1} \cdots \mathcal{G}_{NS}$
as the initial candidate vector $\mathcal{G}$ for the remaining $({1}-\mathcal{F})*\mathbf{N}$ number of function evaluations.
This particle was then updated according to the rule:
\begin{equation}
  \mathcal{G}_{new}(\mathbf{J})=\begin{cases}
    \mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J}), & \text{if $\mathcal{G}_{new}(\mathbf{J})<\mathcal{G}(\mathbf{J})$}.\\
    \mathcal{G}(\mathbf{J}), & \text{otherwise}.
  \end{cases}
\end{equation}
where $\mathbf{J}$ represents the set containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$,
and $\sigma$ is the amplitude of perturbation:
\begin{equation}
	\sigma = \mathbf{R}(\mathcal{MAX} -\mathcal{MIN})
\end{equation}
where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension. The set $\mathbf{J}$ was constructed using a probability function $\mathcal{P}_{j}$ that represented the probability whether a specific dimension $j$ was perturbed or not.  This function is a monotonically decreasing function that decreases with the number of function evaluations. $\mathcal{P}_{j}$ can be any monotonically decreasing function, in our approach we used the following function:
\begin{eqnarray}
	%\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
	\mathcal{P}_{j}&=&{1}-\log(j/ (({1}-\mathcal{FR})*\mathbf{N}))
\end{eqnarray}
%where $\beta$ is the perturbation frequency probability modulator.
Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase. These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$. The decrease of weight function in equation 4, reflection boundary conditions in equation 5, the update function described in equations 6 and 7, the selection probability in equation 8 are the means by which DDS ideas are incorporated into DOPS.
%Equation 5 describing the reflection boundary conditions, equations 6,7 and 8 are the means by which DDS ideas are incorporated into DOPS.
The fraction of evaluations $\mathcal{FR}$  within the swarm phase is based on a switching strategy wherein the switch from swarm phase to DDS phase happens when the error due to the best solution does not drop more than 1\% of the original error, continuously for more than a prescribed number of function evaluations. This allows the solution to quickly jump out of local optima and avoid any convergence issues that are generally associated with swarm based searches.

@@

\subsection*{Performance of DOPS on a model of the human coagulation cascade.}
We compared the performance of DOPS with simulated annealing (SA), differential evolution (DE), and dynamically dimensioned search (DDS) for a model of the human coagulation cascade.
Coagulation is an archetype biochemical network that is highly interconnected, containing both positive and negative feedback (Fig. \ref{fig-coagulation-network}).
In this study, we used the coagulation model of Luan et al \cite{luan2007computationally} which modeled coagulation as a system of coupled non-linear ordinary differential equations.
The Luan model contained 148 parameters and 92 species and has been validated using 21 published datasets.

The biochemistry underlying coagulation, though quite complex, has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model},
and reliable experimental coagulation models have been developed \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}.
Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets.
The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
Thrombin generation involves three phases, initiation, amplification and termination [13,14].
Initiation requires a trigger event, for example vessel injury, which leads to the activation of factor VII (FVIIa).
Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal.
The extrinsic cascade is generally believed to be the main mechanism of thrombinogenesis in the blood [15–17].
Initially, thrombin is produced upon cleavage of prothrombin by fluid phase activated factor X (FXa), which itself has been activated by Tissue Factor/activated factor VII (TF/FVIIa) [10].
Picomolar amounts of thrombin then activate the cofactors factors V and VIII (fV and fVIII) and platelets, leading to the formation of the tenase and prothrombinase complexes on activated platelets.
These complexes amplify the early coagulation signal by further activating FXa, and directly converting prothrombin to thrombin.
There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation. Tissue Factor Pathway Inhibitor (TFPI) inhibits FXa formation catalyzed by TF/FVIIa, while antithrombin III (ATIII) neutralizes several of the proteases generated during coagulation, including thrombin.
Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR), converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of fV/FVa and fVIII/FVIIIa.
Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as APC or ATIII.

@@

To estimate coagulation model parameters, we used data sets from TF-VIIa initiated coagulation in the absence of anticoagulants.
The objective function was a linear combination of two error functions that used data sets representing coagulation initiated with different concentrations of TF-VIIa (5pM, 5nM) \cite{hockin2002model}.
We restricted the number of function evaluations to 4000 for each algorithm we tested, and performed 25 trials of each experiment.
DOPS converged faster and had a lower final error compared to the other algorithms (Fig. \ref{fig-convergence}).
Within the first 1000 function evaluations of DOPS, there was a rapid drop in error.
Approximately between 500-1000 function evaluations a switch to dynamically dimensioned search phase happens (this switch varied from trial to trial since the switch was based on the error from the swarm phase).
Overall at the end of 4000 function evaluations DOPS minimized the error (final objective error is 5.5916e+04) to a greater extent than the other algorithms.
Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Fig.\ref{fig-train}).
The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval.
Subsequently we used these optimal parameters to make model predictions that was compared against completely 'unseen' or untrained experimental data where
coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Fig.\ref{fig-validation}).

\subsection*{Performance of DOPS on benchmark problems and test functions.}
Villaverde and co-workers recently published a set of benchmark biochemical problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}. From a computational cost perspective problems they categorized the problems as most expensive, intermediate and least expensive. We evaluated the performance of our algorithms on a problem from the most expensive and least expensive categories. The first problem (most expensive) is a genome wide kinetic model of \textit{Saccharomyces cerevisiae} with 261 reactions, 262 variables and 1759 parameters (henceforth referred to as problem B1). The reactions were modeled using modular rate law and generalized form of Michaelis-Menten kinetics.   The second problem (henceforth referred to as problem B4, least expensive) is a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters. In both cases pseudo time series data was generated by the authors.  For problem B1, the time series data consisted of 44 observables and for problem B4 the data corresponded to 13 different metabolites. For a detailed description about the model architectures please refer to the work of Villaverde et al. \cite{villaverde2015biopredyn}

@@

@@

\begin{table}[h]
\centering
\caption{Table with optimization settings and results for the coagulation problem, the benchmarks and test functions using DOPS. For each problem the bounds on the parameter vector, the total number of function evaluations, the best initial objective value and the best final objective value are specified. Nominal objective value represents the objective value using the true parameter vector or the nominal parameter vector. The CPU time is the time taken for the problem on a 2.4GHz Intel Xeon Architecture running Matlab 2014b.}
\label{table:benchmark-problems}
\resizebox{\columnwidth}{!}{
\begin{tabular}{|cccccc|}
\hline
                        & Coagulation & B1           & B4          & Ackley 300D & Rastrigin 300D \\
 \hline
Lower Bound             & 0.001.pnom            & 5.pnom       & 5.pnom      & 30          & 5.12           \\
Upper Bound             & 1000.pnom             & 0.2.pnom     & 0.2.pnom    & -15         & -5.12          \\
CPU Time                & 10.0835 hours         & 38.308 hours & 6.2 minutes & 2.8 seconds & 2.6 seconds    \\
Function Evaluations    & 4000                  & 4000         & 4000        & 4000        & 4000           \\
Initial Objective Value & 1.11837e+07           & 1.4275e+07   & 1.8536e+07  & 21.12       & 99.985         \\
Final Objective Value   & 5.5916e+04              & 3.5348e+04     & 38.1306     & 0           & 0              \\
Nominal Objective Value & 4.7785e+06            & 1.0986e+06   & 39.0676     & 0           & 0      \\
\hline
\end{tabular}
}
\end{table}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_3_Errors_convergence.pdf}
\caption{Error convergence rates of the 5 different algorithms on the coagulation model. The objective error is the mean over $N$= 25 trials. DOPS, DDS and SA have the steepest drop in error during first 300 function evaluations. Thereafter the error drop in DDS and SA remains nearly constant whereas DOPS continues to drops further. At the end of 4000 function evaluations DOPS attains the lowest error. The next best estimate using DDS is nearly 3 times greater than the lowest error using DDS.
}\label{fig-convergence}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth,height=0.5\textheight]{./figs/Figure_4_Sim_Train_E1_E5.pdf}
\caption{Model fits on experimental data using DDSMLSPSO. The model parameters were estimated using DOPS. Solid black lines indicate the simulated mean thrombin concentration using parameter vectors from $N$= 25 trials. The grey shaded region represents the 99\% confidence estimate of the mean simulated thrombin concentration. The experimental data is reproduced from the synthetic plasma assays of Mann and co-workers. Thrombin generation is initiated by adding Factor VIIa-TF (5nM - Red and 5pM - Green) to synthetic plasma containing 200 $\mu$mol/L of phospholipid vesicles (PCPS) and a mixture of coagulation factors (II,V,VII,VIII,IX,X and XI) at their mean plasma concentrations.
}\label{fig-train}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth,height=0.5\textheight]{./figs/Figure_6_Ackley_Rast}
\caption{Error convergence on 300 dimensional Ackley function and Rastrigin function. \textbf {(A)} Error convergence of 4 different meta-heuristics on 300 dimensional Ackley function \textbf {(B)} Error convergence of 4 different meta-heuristics on 300 dimensional Rastrigin function
}\label{fig-testfunctions}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth,height=0.7\textheight]{./figs/Figure_B1_B4_params_measuredstates}
\caption{ \textbf {(A)} Difference between nominal and optimal parameters for problem B1: Genome wide kinetic model of \textit{S.cerevisiae} with 1759 unknown parameters. \textbf {(B)} Difference between experimental (measured) data and data simulated with optimal parameters for problem B1: Genome wide kinetic model of \textit{S.cerevisiae} with 1759 unknown parameters. \textbf {(C)} Difference between nominal and optimal parameters for problem B4: Metabolic model of Chinese  Hamster Ovary Cells (CHO) cells with 117 parameters. \textbf {(D)} Difference between experimental (measured) data and data simulated with optimal parameters for problem B4: Metabolic model of Chinese  Hamster Ovary Cells (CHO) cells with 117 parameters.
}\label{fig-benchmark}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth,height=0.7\textheight]{./figs/B1_sims.pdf}
\caption{\textbf {(Data fits for Problem B1)} Pseudo-experimental data (red circles) vs. optimal solution obtained using DOPS (solid blue lines) for the 44 observed states. X axis: time [s]; Y axis: metabolite concentrations [mM].
}\label{fig-sims-b1}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth,height=0.7\textheight]{./figs/FigB4_sims.pdf}
\caption{\textbf {(Data fits for Problem B4)} Pseudo-experimental data (red x) vs. optimal solution obtained using DOPS (solid blue lines) for the 13 observed states. X axis: time [s]; Y axis: metabolite concentrations [mM].
}\label{fig-sims-b4}
\end{figure}

>>>>

\subsection*{Dynamic optimization with particle swarms (DOPS).}
DOPS is a novel meta-heuristic which combines multi-swarm particle swarm methods with the dynamically dimensioned search approach of Shoemaker and colleagues (Fig. \ref{fig-algorithm}).
The goal of DOPS is to estimate optimal or near optimal parameter vectors for high-dimensional biological models within a specified number of function evaluations.
Toward this objective, DOPS begins by using a particle swarm search and then dynamically switches, using an adaptive switching criteria, to the DDS search phase.

\subsubsection*{Phase 1: Swarm phase.}
We began the particle swarm phase of DOPS by randomly initializing a swarm of $\mathcal{K}$-dimensional particles (represented as $z_{i}$),
wherein each particle corresponded to a $\mathcal{K}$-dimensional parameter vector.
After initialization, particles were randomly partitioned into $k$ equal sized sub-swarms $\mathcal{S}_{1},\hdots,\mathcal{S}_{k}$.
Thereafter within each sub-swarm $\mathcal{S}_{k}$, particles were updated according to the rule:
\begin{equation}\label{eqn:update-rule}
	\mathbf{z}_{i,j} = \theta_{1,j-1}\mathbf{z}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{z}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G}_{k} - \mathbf{z}_{i,j-1}\right)
\end{equation}
where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ were adjustable parameters, $\mathcal{L}_{i}$ denotes the best solution found by particle $i$ within sub-swarm $k$ for function evaluation $1\rightarrow j-1$, and
$\mathcal{G}_{k}$ denotes the best solution found over all particles within sub-swarm $\mathcal{S}_{k}$.
The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model parameters ($\mathcal{K}\times{1}$).
Equation (3) is similar to the general particle swarm update rule, however, it does not contain velocity terms.
In DOPS the parameter $\theta_{1,j-1}$ depends upon the function evaluations and is updated according to:
\begin{eqnarray}
	\mathbf \theta_{1,j}&=&\frac{(\mathcal{N}-{j})*(\mathbf{w}_{max}-\mathbf{w}_{min}))}{(\mathcal{N}-{1})} + \mathbf{w}_{min}
\end{eqnarray}
where $\mathcal{N}$ represents the total number of function evaluations, $\mathbf{w}_{max}$ and $\mathbf{w}_{min}$ are the maximum and minimum inertia weights, respectively.
While updating the particles, we made sure all dimensions of the solution represented by the particle were within bounds using a set of reflection boundary conditions:

@@

After every $\mathcal{M}$ function evaluations, particles were randomly redistributed to a new sub-swarm, and updated according to Eqn. \eqref{eqn:update-rule}.
This process continued for $\mathcal{F}*\mathcal{N}$ functions evaluations, where $\mathcal{F}$ is the fraction of evaluations in the particle swarm phase of DOPS.
If the simulation error stagnates for example, does not change by more than 1\% for a specified number of evaluations, the swarm phase is terminated and DOPS switches to
exploring parameter space using a DDS approach.

\subsubsection*{Phase 2: DDS phase.}
At the conclusion of the swarm phase, the overall best particle, $\mathcal{G}_{k}$, over the k sub-swarms was to initalize the DDS phase.
DOPS takes $\left(1-\mathcal{F}\right)*\mathcal{N}$ function evaluations in the DDS phase and then terminates the search.
For the DDS phase, the best parameter estimate was updated using the rule:
\begin{equation}
  \mathcal{G}_{new}(\mathbf{J})=\begin{cases}
    \mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J}), & \text{if $\mathcal{G}_{new}(\mathbf{J})<\mathcal{G}(\mathbf{J})$}.\\
    \mathcal{G}(\mathbf{J}), & \text{otherwise}.
  \end{cases}
\end{equation}
where $\mathbf{J}$ represents the set containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$,
and $\sigma$ denotes the perturbation amplitude:
\begin{equation}
	\sigma = \mathbf{R}(\mathcal{MAX} -\mathcal{MIN})
\end{equation}
where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension. The set $\mathbf{J}$ was constructed using a probability function $\mathcal{P}_{j}$ that represented the probability whether a specific dimension $j$ was perturbed or not.  T
his function is a monotonically decreasing function that decreases with the number of function evaluations. $\mathcal{P}_{j}$ can be any monotonically decreasing function, in our approach we used the following function:
\begin{eqnarray}
	%\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
	\mathcal{P}_{j}&=&{1}-\log(j/ (({1}-\mathcal{FR})*\mathbf{N}))
\end{eqnarray}
%where $\beta$ is the perturbation frequency probability modulator.
Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase. These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$. The decrease of weight function in equation 4, reflection boundary conditions in equation 5, the update function described in equations 6 and 7, the selection probability in equation 8 are the means by which DDS ideas are incorporated into DOPS.
%Equation 5 describing the reflection boundary conditions, equations 6,7 and 8 are the means by which DDS ideas are incorporated into DOPS.
The fraction of evaluations $\mathcal{FR}$  within the swarm phase is based on a switching strategy wherein the switch from swarm phase to DDS phase happens when the error due to the best solution does not drop more than 1\% of the original error, continuously for more than a prescribed number of function evaluations. This allows the solution to quickly jump out of local optima and avoid any convergence issues that are generally associated with swarm based searches.

@@

\subsection*{Performance of DOPS on a model of the human coagulation cascade.}
We compared the performance of DOPS with simulated annealing (SA), differential evolution (DE), and dynamically dimensioned search (DDS) for a model of the human coagulation cascade.
Coagulation is an archetype biochemical network that is highly interconnected, containing both positive and negative feedback (Fig. \ref{fig-coagulation-network}).
The biochemistry of coagulation, though complex, has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model},
and reliable experimental protocols have been developed to interrogate the system \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}.
Thus, the human coagulation cascade is an ideal test case;
coagulation is challenging because it contains both fast and slow dynamics, but also accessible because of the availability of comprehensive data sets for model identification and validation.
In this study, we used the coagulation model of Luan et al \cite{luan2007computationally}. Luan and coworkers modeled coagulation as a system of coupled non-linear ordinary differential equations.
The Luan model contained 148 parameters and 92 species and has been validated using 21 published datasets.

Coagulation is mediated by a family proteases in the circulation, called factors and a key group of blood cells, called platelets.
The central process in coagulation is the conversion of prothrombin (fII), an inactive coagulation factor, to the master protease thrombin (FIIa).
Thrombin generation involves three phases, initiation, amplification and termination [13,14].
Initiation requires a trigger event, for example a vessel injury which exposes tissue factor (TF), which leads to the activation of factor VII (FVIIa) and the formation of the TF/FVIIa complex.
Two converging pathways, the extrinsic and intrinsic cascades, then process and amplify this initial coagulation signal.
There are several control points in the cascade that inhibit thrombin formation, and eventually terminate thrombin generation.
Tissue Factor Pathway Inhibitor (TFPI) inhibits upstream activation events, while antithrombin III (ATIII) neutralizes several of the proteases generated during coagulation, including thrombin.
Thrombin itself also inadvertently plays a role in its own inhibition; thrombin, through interaction with thrombomodulin, protein C and endothelial cell protein C receptor (EPCR), converts protein C to activated protein C (APC) which attenuates the coagulation response by proteolytic cleavage of amplification complexes.
Termination occurs after either prothrombin is consumed, or thrombin formation is neutralized by inhibitors such as APC or ATIII.

@@

To estimate the coagulation model parameters, we used data sets from TF/VIIa initiated coagulation in the absence of anticoagulants.
The objective function was a linear combination of two error functions, representing coagulation initiated with different concentrations of TF-VIIa (5pM, 5nM) \cite{hockin2002model}.
We restricted the number of function evaluations to $\mathcal{N}$ = 4000 for each algorithm we tested, and performed 25 trials of each experiment to collect average performance data (Table ZZ).DOPS converged faster and had a lower final error compared to the other algorithms (Fig. \ref{fig-convergence}).
Within the first 1000 function evaluations of DOPS, there was a rapid drop in error.
Approximately between 500-1000 function evaluations a switch to dynamically dimensioned search phase happens (this switch varied from trial to trial since the switch was based on the error from the swarm phase). Overall at the end of 4000 function evaluations DOPS minimized the error (final objective error is 5.5916e+04) to a greater extent than the other algorithms.

Using the parameters obtained at the end of 4000 function evaluations we examined the 'fits' between models predictions and experimental data (Fig.\ref{fig-train}).
The solid lines represent the mean value of prediction over 25 trials and the shaded region represents the 99\% confidence interval.
Subsequently we used these optimal parameters to make model predictions that was compared against completely 'unseen' or untrained experimental data where
coagulation was initiated with 500pM,50pM,10pM concentrations of TF-VIIa respectively (Fig.\ref{fig-validation}).

\subsection*{Performance of DOPS on benchmark problems.}
Villaverde and co-workers recently published a set of benchmark biochemical problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}. From a computational cost perspective problems they categorized the problems as most expensive, intermediate and least expensive. We evaluated the performance of our algorithms on a problem from the most expensive and least expensive categories. The first problem (most expensive) is a genome wide kinetic model of \textit{Saccharomyces cerevisiae} with 261 reactions, 262 variables and 1759 parameters (henceforth referred to as problem B1). The reactions were modeled using modular rate law and generalized form of Michaelis-Menten kinetics.  The second problem (henceforth referred to as problem B4, least expensive) is a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters. In both cases pseudo time series data was generated by the authors.  For problem B1, the time series data consisted of 44 observables and for problem B4 the data corresponded to 13 different metabolites. For a detailed description about the model architectures please refer to the work of Villaverde et al. \cite{villaverde2015biopredyn}

@@

\subsection*{Influence of the number of sub-swarms on DOPS performance.}

@@

\begin{table}[h]
\centering
\caption{Table with optimization settings and results for the coagulation problem, the benchmarks and test functions using DOPS. For each problem the bounds on the parameter vector, the total number of function evaluations, the best initial objective value and the best final objective value are specified. Nominal objective value represents the objective value using the true parameter vector or the nominal parameter vector. The CPU time is the time taken for the problem on a 2.4GHz Intel Xeon Architecture running Matlab 2014b.}
\label{table:benchmark-problems}
\resizebox{\columnwidth}{!}{
\begin{tabular}{r|lllll}
\hline\\
                        & Coagulation 					& B1           	& B4          & Ackley 			& Rastrigin \\\\
\hline\\
Evaluations   					& 4000                  & 4000         	& 4000        & 4000        & 4000        \\
Lower Bound             & 0.001.pnom            & 5.pnom       	& 5.pnom      & 30          & 5.12        \\
Upper Bound             & 1000.pnom             & 0.2.pnom     	& 0.2.pnom    & -15         & -5.12       \\
CPU Time                & 10.1 hrs         			& 38.3 hrs 			& 6.2 min 		& 2.8 s 			& 2.6 s    		\\\\

Scaled initial error 		& 1.0           				& 1.0   				& 1.0  				& 1.0       	& 1.0         \\
Scaled final error  		& $<$ 0.01           		& $<$ 0.01   		& $<$ 0.01    & $<$ 0.01   	& $<$ 0.01    \\
Scaled nominal error 		& 0.42            			& 0.1   				& $<$ 0.01    & 0           & 0      			\\

\hline
\end{tabular}
}
\end{table}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure_3_Errors_convergence_v2.pdf}
\caption{Error convergence rates of the 5 different algorithms on the coagulation model. The objective error is the mean over $N$= 25 trials. DOPS, DDS and SA have the steepest drop in error during first 300 function evaluations. Thereafter the error drop in DDS and SA remains nearly constant whereas DOPS continues to drops further. At the end of 4000 function evaluations DOPS attains the lowest error. The next best estimate using DDS is nearly 3 times greater than the lowest error using DDS.
}\label{fig-convergence}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure_4_Sim_Train_E1_E5.pdf}
\caption{Model fits on experimental data using DDSMLSPSO. The model parameters were estimated using DOPS. Solid black lines indicate the simulated mean thrombin concentration using parameter vectors from $N$= 25 trials. The grey shaded region represents the 99\% confidence estimate of the mean simulated thrombin concentration. The experimental data is reproduced from the synthetic plasma assays of Mann and co-workers. Thrombin generation is initiated by adding Factor VIIa-TF (5nM - Red and 5pM - Green) to synthetic plasma containing 200 $\mu$mol/L of phospholipid vesicles (PCPS) and a mixture of coagulation factors (II,V,VII,VIII,IX,X and XI) at their mean plasma concentrations.
}\label{fig-train}
\end{figure}

@@

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./figs/Figure_6_Ackley_Rast}
\caption{Error convergence on 300 dimensional Ackley function and Rastrigin function. \textbf {(A)} Error convergence of 4 different meta-heuristics on 300 dimensional Ackley function \textbf {(B)} Error convergence of 4 different meta-heuristics on 300 dimensional Rastrigin function
}\label{fig-testfunctions}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./figs/Figure_B1_B4_params_measuredstates}
\caption{ \textbf {(A)} Difference between nominal and optimal parameters for problem B1: Genome wide kinetic model of \textit{S.cerevisiae} with 1759 unknown parameters. \textbf {(B)} Difference between experimental (measured) data and data simulated with optimal parameters for problem B1: Genome wide kinetic model of \textit{S.cerevisiae} with 1759 unknown parameters. \textbf {(C)} Difference between nominal and optimal parameters for problem B4: Metabolic model of Chinese  Hamster Ovary Cells (CHO) cells with 117 parameters. \textbf {(D)} Difference between experimental (measured) data and data simulated with optimal parameters for problem B4: Metabolic model of Chinese  Hamster Ovary Cells (CHO) cells with 117 parameters.
}\label{fig-benchmark}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./figs/B1_sims.pdf}
\caption{\textbf {(Data fits for Problem B1)} Pseudo-experimental data (red circles) vs. optimal solution obtained using DOPS (solid blue lines) for the 44 observed states. X axis: time [s]; Y axis: metabolite concentrations [mM].
}\label{fig-sims-b1}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=1.00\textwidth]{./figs/FigB4_sims.pdf}
\caption{\textbf {(Data fits for Problem B4)} Pseudo-experimental data (red x) vs. optimal solution obtained using DOPS (solid blue lines) for the 13 observed states. X axis: time [s]; Y axis: metabolite concentrations [mM].
}\label{fig-sims-b4}
\end{figure}
